# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
```{r}
library("graphics")
library("timeDate")
library("lattice")
activity <- read.csv("~/GitHub/RepData_PeerAssessment1/activity/activity.csv")
# View(activity)
# aggregate the number of steps by date
s1 = aggregate(steps ~ date, data = activity, FUN=sum)
aggrByInterval = aggregate(steps ~ interval, data = activity, FUN=mean)
```


## What is mean total number of steps taken per day?
#### 1. Make a histogram of the total number of steps taken each day
```{r echo = TRUE}
#1. create a histogram of the total number of steps taken eacn day
hist(s1$steps, xlab="Number of Steps taken per day", main="Steps Taken Each Day")

```

#### 2. Calculate the mean and median total number of steps taken per day
```{r}
mean1 = sapply(s1["steps"], mean, na.rm=TRUE)
print (paste("mean of steps taken = ", mean1))
#
median1 = sapply(s1["steps"], median, na.rm=TRUE)
print (paste("median of steps taken=", median1))

```

## What is the average daily activity pattern?
#### 1. Make a time series plot of the 5-minutes interval and the avg number of steps taken 
```{r}
plot(aggrByInterval, type="l")
```
#### 2. Which 5-minute interval on avg across all the days in the dataset, contains the maximum number of steps?
```{r}
print(paste("Interval with the max number of steps = ", aggrByInterval[aggrByInterval["steps"] == max(aggrByInterval["steps"]),1]))

```

## Imputing missing values
#### 1. Calculate and report the total number of missing values in the dataset ( i.e the total number of rows with NA)

```{r}
ok = complete.cases(activity)
print(paste("Number of rows with missing values = ", sum(!ok)))

# just to confirm there are no rows with missing date or interval 
sum(is.na(activity$date))
sum(is.na(activity$interval))
```
#### 2. the strategy to fill the missing values is to use the mean of steps taken for that interval

#### 3. create a new dataset that is equal to the original dataset but with missing data filled in using the strategy from step 2

```{r}
completeActivity = activity
for(i in 1:nrow(activity)){
  if (is.na(activity[i,"steps"])) {
    int1 = activity[i,"interval"]
    completeActivity[i,"steps"] = aggrByInterval[which(aggrByInterval["interval"] == int1),"steps"]
  }
}
# confirm that there no more rows with missing values
ok = complete.cases(completeActivity)
print(paste("Number of rows with missing values = ", sum(!ok)))

```
#### 4. Make a histogram of the number of steps taken each day and calculate the mean and median total number of steps taken per day.
```{r}
# aggregate the number of steps by date
s2 = aggregate(steps ~ date, data = completeActivity, FUN=sum)
hist(s2$steps, xlab="Number of Steps taken per day", main ="Steps Taken Each Day")

#aggrByInter2 = aggregate(steps ~ interval, data = completeActivity, FUN=mean)
mean2 = sapply(s2["steps"], mean, na.rm=TRUE)
print (paste("mean of steps taken = ", mean2))
#
median2 = sapply(s2["steps"], median, na.rm=TRUE)
print (paste("median of steps taken=", median2))

print (paste("Mean with missing values =", mean1, " vs mean with missing values replaced by the mean for respective interval = ", mean2))

print (paste("Median with missing values =", median1, " vs mean with missing values replaced by the median for respective interval = ", median2))

```
#### Since the data is missing for an entire day, replacing the missing values with the respective mean for the interval did not impact the mean or median for the day

## Are there differences in activity patterns between weekdays and weekends?
#### 1. Create a new varable called weekday in the completeActivity dataset
```{r}
completeActivity$weekday = isWeekday(completeActivity$date)
```
```{r}
aggrByIntervalWeekday = aggregate(steps ~ interval + weekday, data = completeActivity, FUN=mean) 
```

#### 2. Make a panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday day or weekend days 
```{r}
aggrByIntervalWeekday$weekend = ifelse(aggrByIntervalWeekday$weekday, "Weekday","Weekend")
xyplot(steps ~ interval | weekend, data=aggrByIntervalWeekday, type = "l", layout=c(1,2))
```